A Plasma Torus Around a Young Low-Mass Star

### 2025-04-07 paper/ms.tex
update ms

\begin{abstract}
\normalfont
Roughly one percent of red dwarfs younger than 100 million years show
structured, periodic optical light curves suggestive of transiting
opaque material that corotates with the star
\cite{Rebull2016,Stauffer2017,Rebull2018,Bouma2024}.  However, the
composition, origin, and even the existence of this material are
uncertain. The main alternative hypothesis is that these complex
periodic variables (CPVs) are explained by complex distributions of
bright or dark regions on the stellar surfaces \cite{Koen2021}.
Here, we present time-series spectroscopy and photometry of a
rapidly-rotating ($P$=3.9\,hr) CPV, TIC~141146667. The spectra show
sinusoidal time-varying Balmer emission at twice to four times the
star's equatorial velocity, directly demonstrating for the first
time the existence of corotating circumstellar plasma around a CPV.
Given that long-lived condensations of cool ($10^4$ K) plasma can
persist in the hot ($10^6$ K) coronae of stars with a wide range of
masses \cite{CollierCameron1989,Townsend2005,Dunstone2006,Petit2013,Waugh2022,Daley-Yates2024},
these data support the idea that around the lowest-mass stars, such
condensations can sustain an optically thick component; the origin
of this optically thick material and its microphysical opacity remain
unclear.
\end{abstract}

@@

M dwarfs, stars with masses below about half that of the Sun, are the
only type of star to offer near-term prospects for detecting the
atmospheres of rocky exoplanets with surface water \cite{NAP26141}.
Community investment with JWST has proceeded accordingly
\cite{Redfield2024,TRAPPIST1JWSTCommunityInitiative2024}.  How an M
dwarf's evolution influences its planets---especially the retention of
their atmospheres--- has in turn become a major theme in
exoplanet and stellar astrophysics.  Previous work has established
that most M dwarfs host close-in planets \cite{Dressing2015} that 
on average are subject to long circumstellar disk lifetimes
\cite{Ribas2015}, to high doses of UV radiation \cite{France2013},
and to a high incidence of flares and coronal mass ejections
\cite{Gunther2020}.  However, despite extensive work in these areas,
the plasma and magnetospheric environments that bathe young, close-in
exoplanets remain challenging to quantify.
Understanding these environments is crucial because they directly
impact atmospheric retention and habitability of close-in planets.

One glaring example of our current ignorance is the complex periodic
variables (CPVs).  While Figure~\ref{fig:lc} highlights the main
object of interest in this article, over one hundred analogous objects
have now been found by K2 and TESS
\cite{Rebull2016,Stauffer2017,Rebull2018,Zhan2019,Rebull2020,Bouma2024}.
These CPVs are phenomenologically identified based on their
structured, periodic optical light curves; most are M dwarfs with
rotation periods shorter than two days.  Within current sensitivity
limits, none host disks \cite{Stauffer2017,Bouma2024}.
However, $\approx$3\% of stars a few million years old show complex
photometric behavior, an observed fraction which decreases to $\approx$0.3\% by
$\approx$150\,Myr \cite{Rebull2020}.

@@

The dearth of evidence for circumstellar material around CPVs is
surprising given that separate studies of young stars have, for
decades, reported that stellar coronae contain both hot ($10^6$ K) and
cool ($10^4$ K) plasma. In particular, time-series spectra of stars
with a wide range of masses have shown periodic high-velocity
absorption and emission in Balmer lines such as H$\alpha$, interpreted
as long-lived, corotating clumps of cool plasma
\cite{CollierCameron1989,CollierCameron1992,Barnes2000,Donati2000,Dunstone2006,Skelly2008,Leitzinger2016,Cang2021}.
Such clumps are thought to be forced into corotation by the magnetic
field, and the exact geometry of where the plasma can accumulate is
dictated by the field's topology.  For instance, a magnetic dipole field
tilted with respect to the stellar spin axis
yields accumulations in a warped torus geometry
\cite{Townsend2005}, whereas in the limit of a single strong field
line, accumulation occurs at the line's apex, furthest from the star
\cite{Waugh2022}.  To date, none of these spectroscopic variables have
shown any photometric anomalies \cite{Bouma2024}, leaving open the
issue of whether they are related to CPVs.

In this study, we present the first spectroscopic detection of
corotating clumps of cool plasma around a CPV.  We identified
TIC~141146667 in previous work \cite{Bouma2024} by searching the TESS
two-minute data \cite{Ricker2015} for stars showing highly structured,
periodic light curves.  We selected the star for spectroscopy because
its brightness and rotation rate enable efficiently measuring
variability in its line profile.  We observed it for five hours on
17~February~2024 (UT) using the High Resolution Echelle Spectrometer
(HIRES \cite{vogt_hires_1994}) on the 10\,m Keck I telescope.  TESS
observed the star over a four week interval from 30~January~2024 to
26~February~2024 with a duty cycle of 77\%.  During the HIRES
observations, TESS was performing a data downlink; TESS data
collection resumed 12 hours (three rotation cycles) after the spectra
were acquired.  Extended Data Figure~\ref{fig:fulllc} shows the
detailed photometric behavior of the star near the epoch of
observation; the general photometric morphology is similar before and
after the data gap.

@@

% ** keck/hires: emission beyond v/veq>1
The spectroscopy shows emission beyond the star's equatorial velocity
($v_{\rm eq}$=130\,\kms).  There are at least two distinct emission
components, separated by half a cycle in phase.  The first component
has clearer sinusoidal behaviour in time and is double-peaked, with peak
semi-amplitudes of $K_1$=2.1\,$v_{\rm eq}$ and 2.7\,$v_{\rm eq}$.
There is significant non-periodic variability in the emissivity of
this double-peaked component: the flux excess begins with an amplitude
of 70\% above the continuum early in the observation sequence, and
diminishes to 10\% by its end.  The component 180$^\circ$ opposite in
phase is detected only from $\phi$=0.2-1.0.  From $\phi$=0.2-0.5, this
latter component appears connected to the star in velocity space.
While its peak semi-amplitude of $K_1$=3.9\,$v_{\rm eq}$ is achieved
at both $\phi$=0.25 and 0.75, its amplitude similarly decreases from a
60\% excess over the continuum at the beginning of the observation
sequence to a 10\% excess by its end.  The apparent period for all
of these emission components is consistent with the photometric
\periodhr\ hour period.

These sinusoidal emission features require circumstellar clumps of
partially-ionized hydrogen to corotate with the star.  Based on the
observed sinusoidal period and peak velocity, this material's motion
is not controlled by gravitational attraction to the star; it can only be
explained by plasma being dragged along with the rotating stellar
magnetic field.  The velocity semi-amplitude of the sinusoids gives
the distance of the clumps from the stellar surface:
2.1-2.7\,$R_\star$ for the closer clump, and 3.9\,$R_\star$ for the
other.   These clumps transit in front of the star when passing from
negative to positive velocity.  This implies that the spectroscopic
transit of the 2.1-2.7\,$R_\star$ clump occurs simultaneously with the
sharp photometric spike visible in the TESS data.

% * keck/hires: behavior at v/veq<1
The H$\alpha$ line core is more complex.  At $|\Delta v /
v_{\rm eq}|<1$, most observed H$\alpha$ photons come from the star's
chromosphere; circumstellar material might also modulate the
line profile.
Figure~\ref{fig:spec}e suggests line core variability caused by both
bright and dark regions on the star's surface, superposed
with smaller-amplitude variability from the transiting circumstellar
material.
For instance, from $\phi$=0-0.3, the double-peaked emission feature
is visible when it is being viewed both on and off-limb.
However, the large emission feature that crosses the star from
$\phi$=0.4-0.9 cannot be circumstellar in origin; it emits at an
amplitude greater than that observed from the circumstellar components,
and it crosses the stellar velocity surface too slowly to be
circumstellar.
It must therefore originate from a chromospherically bright region
on the star's surface.
While the origin of the other bright and dark streaks passing across the
line core are more ambiguous,
a final exercise to quantify the behavior of this region is shown in
Figure~\ref{fig:spec}d, where $f_{\rm H\alpha\ core}$ denotes the
summed flux at $|\Delta v / v_{\rm eq}|<1$.  This panel shows that
changes in the line core flux correlate with the broadband variability
throughout most of the light curve, except near $\phi$$\approx$0.5,
corresponding to the transit of the 3.9\,$R_\star$ clump and the
occultation of the lower-velocity clump.

\begin{table*}
\small
\setlength{\tabcolsep}{2pt}
\centering
\caption{Selected System Parameters for TIC~141146667}
\label{tab:params}
\begin{tabular}{llcc}
\hline \hline
Parameter & Description & Value & Source\\
\hline 
%
$T_{\rm eff}$\dotfill                   & Effective Temperature (K) \hspace{9pt}\dotfill                 & 2972 $\pm$ 40    & 1 \\
%
$R_\star$\dotfill                       & Stellar radius ($R_\odot$)\dotfill                             & 0.42$\pm$0.02    & 1 \\
%
Age                                     & Stellar age range (Myr)\dotfill                                & 35-150           & 2 \\
%
$M_\star$\dotfill                       & Stellar mass ($M_\odot$)\dotfill                               & 0.22$\pm$0.02    & 3 \\
%
$\gamma$\dotfill                        & Systemic radial velocity (\kms)\dotfill                        & 0.61 $\pm$ 1.47  & 4 \\
%
Spec. Type\dotfill                      & Spectral Type\dotfill                                          & M5.5Ve           & 4 \\
%
$P_{\rm rot}$\dotfill                   & Photometric rotation period (hr)\dotfill                       & $3.930\pm 0.001$ & 5 \\
%
$v_{\rm eq}$\dotfill		                & Equatorial velocity \dotfill                                   &  130$\pm$4       & 6 \\
                                        & \hspace{3pt} ($2\pi R_\star/P_{\rm rot}$) (\kms)	             &                      \\
%
$v_{\rm eq}\sin{i_\star}$\dotfill		    & Projected rotational\dotfill                                   &  138$\pm$8       & 4 \\
                                        & \hspace{3pt} velocity (\kms)	                                 &                      \\
%
$v_{\rm break}$\dotfill		              & Breakup velocity \dotfill                                      &  316$\pm$16      & 6 \\
                                        & \hspace{3pt} ($G M_\star / R_\star$)$^{1/2}$ (\kms)	           &                      \\
%
$i_\star$\dotfill                       & Stellar inclination\dotfill                                    & 	$>$63           & 4 \\
                                        & \hspace{3pt}  2$\sigma$ lower limit (deg)	                     &                      \\
%
$d$\dotfill                             & Distance (pc)\dotfill                                          & $57.54 \pm 0.09$ & 7 \\
%
$R_{\rm c}$\dotfill		                  & Keplerian corotation\dotfill                                   & $1.82 \pm 0.10$  & 6 \\
                                        & \hspace{3pt} radius ($R_\star$)	                               &                      \\
%
$r_0$\dotfill                           & Inner clump (0) orbital radius ($R_\star$)\hspace{9pt}\dotfill         &  2.07$\pm$0.04                 & 4 \\
$r_1$\dotfill                           & Inner clump (1) orbital radius ($R_\star$)\hspace{9pt}\dotfill         &  2.88$\pm$0.10                 & 4 \\
$r_2$\dotfill                           & Outer clump orbital radius ($R_\star$)\hspace{9pt}\dotfill         &  4.16$\pm$0.56                 & 4 \\
%
$\langle$EW$_{\rm H\alpha}$$\rangle$    & Time-averaged H$\alpha$ line core                              &  7.2 $\pm$ 0.2   & 4 \\ 
                                        & \hspace{3pt} equivalent width (\AA)	                           &                      \\
\hline
\end{tabular}
\begin{flushleft}
\footnotesize{ \textsc{NOTE}---
Provenances are:
1: SED fit \cite{Bouma2024}.
2: Gaia DR3 photometry shows the star is on the pre-main sequence,
   while the spectrum lacks lithium (see \S~\ref{subsec:stparams}).
3: PARSEC v1.2S \cite{Chen2014}.
4: Keck/HIRES (see \S~\ref{subsec:halpha}).
5: TESS light curve.
6: Derived quantity.
7: Gaia DR3 geometric \cite{GaiaDR3}.
}
\end{flushleft}
\vspace{-0.5cm}
\end{table*}

\section{Discussion}
% argue: how does this clarify wider phenomenology?  what general
% conclusions can be drawn?

Spectra of magnetically-active, rapidly rotating stars with a wide range
of masses have been known to exhibit both sinusoidal emission features
in time
\cite{Donati2000,Townsend2005,Dunstone2006,Skelly2008} and
transient absorption features
\cite{CollierCameron1989,CollierCameron1992,Cang2020} similar to
Figure~\ref{fig:spec}.  No such stars were previously known to show
complex light curves \cite{Bouma2024}.   One interpretation for
such spectroscopic variability comes from an analogy to quiescent solar
prominences, which are cool condensations of plasma in the solar corona
that can last days to weeks \cite{VialEngvold2015}.  These
condensations fall back to the Sun's surface because gravity is
stronger than any magnetic tension or centrifugal force capable of
sustaining them.  However for stars with magnetospheric radii $R_{\rm
m}$ that exceed their corotation radii $R_{\rm c}$, the effective
potential experienced by a charged particle can have a local minimum
outside $R_{\rm c}$, enabling the material to accumulate over much
longer timescales \cite{Petit2013,Daley-Yates2024}.  Generally
speaking, such material need neither transit nor be opaque in
broadband optical light.
For instance, the H$\alpha$ emission clump at 3.9\,$R_\star$ shows no 
photometric signal in the February 2024 photometry.

Our Keck/HIRES observations are the first reported time-series spectra
of a CPV, and they show that corotating circumstellar plasma clumps
are present around at least one such star.  This observation rules out
a ``starspot-only'' origin for the CPVs, \cite{Koen2021} because
such scenarios have no means of explaining spectroscopic emission
beyond the stellar disk.  Similarly, scenarios in which the
circumstellar material is made only of dust are also ruled out.  While
dust may be present, to explain the H$\alpha$ emission the clumps must
include plasma with a significant population of hydrogen atoms in the
$n$=3 excited state, where $n$ denotes the principal quantum number.
More quantitatively, the observed H$\alpha$ luminosity suggests
characteristic densities and masses for the gaseous component of these
clumps of $n_{\rm H} \sim 10^{11}$-$10^{12}$\,cm$^{-3}$ and $M_{\rm
gas} \sim 10^{17}$\,g (see Section~\ref{subsec:gas}).  Dust, if
present, is independently constrained to have a total mass
$10^{15}\,{\rm g} < M_{\rm dust} < 10^{17}\,{\rm g}$ (see
Section~\ref{subsec:dust}).  This circumstellar material could
originate either from the star or from an external source.  Plausible
external sources include an undetected old disk, comets, or a
close-in exoplanet.  This latter scenario would make CPVs extrasolar
analogs of the Jupiter-Io plasma torus (e.g.~Ref.~\cite{Bagenal1981}).

The other potential analog for the CPVs are the $\sigma$~Ori~E
variables, a rare subset of B stars with radiatively-driven winds which
accumulate into warped plasma tori \cite{Townsend2005,Townsend2008}.
These tori tend to have dense antipodal accumulations of plasma
sculpted by tilted-dipole magnetic fields, and the transits of these
clumps produce broadband photometric variability through bound-free
scattering \cite{Townsend2005} and Thomson scattering
\cite{Berry2022}.  
The plasma accumulates at antipodes 180$^\circ$ apart because the
deepest local minima in the effective potential exist along the line
of intersection between the rotational and magnetic equators (see
Equation~22 of \cite{Townsend2005}).
For $\sigma$~Ori~E and almost
all of its analogs, the result is ``simple'' light curves that 
resemble those of eclipsing binaries, and time-dynamic H$\alpha$
spectra similar to those in Figure~\ref{fig:spec}
\cite{Townsend2005,Townsend2008}.  The two known exceptions, HD~37776
and HD~64740, show complex light curves resembling CPVs
\cite{Mikulasek2020,Bouma2024} and have spectropolarimetric magnetic
field maps indicating strong contributions from higher-order magnetic
moments \cite{Kochukhov2011,Shultz2018}.  There are two implications:
first, the complexity of CPVs may be a direct consequence of magnetic
fields with highly multipolar contributions.  Second, CPVs can be a
source of astrophysical false positives in photometric classification
of eclipsing binaries and transiting exoplanets around young pre-main
sequence M dwarfs \cite{Johns-Krull2016,Bouma2020}.

@@

The material's composition -- either pure plasma or dusty
plasma -- can be clarified by time-series optical and infrared
spectrophotometry.  While observations of CPVs in the optical 
suggest a chromaticity consistent with dust
\cite{Tanimoto2020,Gunther2022,Koen2023}, a gray opacity source such
as electron scattering in a plasma transiting over starspots might
also produce chromatic features \cite{Rackham2018}.  The composition
and size distribution of any dust that is present could be determined
by measuring the extinction curve for one or more CPVs from
1-20\,$\mu$m.  Composition and size distributions similar to debris
from rocky bodies seen around white dwarfs \cite{Reach2009} would
indicate a rocky-body origin.  Compositions and sizes similar to the
interstellar medium could be indicative of condensed dust in an M
dwarf wind, similar to that formed around more evolved stars
\cite{Marigo2008}.

@@

\item[Author Contributions] LGB is responsible for all aspects of this
  work, including conceptualization, data analysis, interpretation,
  manuscript preparation, and revision.

\item[Data Availability] The TESS two-minute cadence data used in this
  work are publicly available from the Mikulski Archive for Space
  Telescopes (MAST; \url{https://mast.stsci.edu}). The Keck/HIRES
  spectra analyzed here are accessible via the Keck Observatory
  Archive (KOA; \url{https://koa.ipac.caltech.edu}).

>>>>

\begin{abstract}
\normalfont
Roughly one percent of red dwarfs younger than 100 million years show
structured, periodic optical light curves suggestive of transiting
opaque material that corotates with the star
\cite{Rebull2016,Stauffer2017,Rebull2018,Bouma2024}.  However, the
composition, origin, and even the existence of this material are
uncertain. The main alternative hypothesis is that these complex
periodic variables (CPVs) are explained by complex distributions of
bright or dark regions on the stellar surfaces \cite{Koen2021}.
Here, we present time-series spectroscopy and photometry of a
rapidly-rotating ($P$=3.9\,hr) CPV, TIC~141146667. The spectra show
sinusoidal time-varying Balmer emission at twice to four times the
star's equatorial velocity, directly demonstrating for the first
time the existence of corotating clumps of circumstellar plasma around a CPV.
Given that cool ($10^4$ K) plasma can persist in the hot ($10^6$ K)
coronae of rapidly-rotating stars with a wide range of masses
\cite{CollierCameron1989,Townsend2005,Dunstone2006,Petit2013,Waugh2022,Daley-Yates2024},
these data support the idea that the magnetospheres of the
lowest-mass stars can support long-lived, optically thick
clumps of material that explain the sharp photometric features seen in
CPVs.  The origin of this optically thick material and its
microphysical opacity remain unclear.
\end{abstract}

@@

M dwarfs, stars with masses below about half that of the Sun, are the
only type of star to offer near-term prospects for detecting the
atmospheres of rocky exoplanets with surface water \cite{NAP26141}.
Community investment with JWST has proceeded accordingly
\cite{Redfield2024,TRAPPIST1JWSTCommunityInitiative2024}.  How an M
dwarf's evolution influences its planets---especially the retention of
their atmospheres--- has in turn become a major theme in
exoplanet and stellar astrophysics.  Previous work has established
that most M dwarfs host close-in planets \cite{Dressing2015} that 
on average are subject to long circumstellar disk lifetimes
\cite{Ribas2015}, to high doses of UV radiation \cite{France2016},
and to a high incidence of flares and coronal mass ejections
\cite{Gunther2020}.  However, despite extensive work in these areas,
the plasma and magnetospheric environments that bathe young, close-in
exoplanets remain challenging to quantify.
Understanding these environments is crucial because they directly
impact atmospheric retention and habitability of close-in planets.

One glaring example of our current ignorance is the complex periodic
variables (CPVs).  While Figure~\ref{fig:lc} highlights the main
object of interest in this article, over one hundred analogous objects
have now been found by K2 and TESS
\cite{Rebull2016,Stauffer2017,Rebull2018,Zhan2019,Rebull2020,Bouma2024}.
These CPVs are phenomenologically identified based on their
structured, periodic optical light curves; most are M dwarfs with
rotation periods shorter than two days.  Within current sensitivity
limits, none host disks \cite{Stauffer2017,Bouma2024}.
However, $\approx$3\% of stars a few million years old show this complex
photometric behavior, an observed fraction which decreases to $\approx$0.3\% by
$\approx$150\,Myr \cite{Rebull2020}.
CPVs are a known source of astrophysical false positives in searches for
transiting exoplanets around young M dwarfs
\cite{vanEyken2012,Johns-Krull2016,Bouma2020}.

@@

The dearth of evidence for circumstellar material around CPVs is
surprising given that separate studies of young stars have, for
decades, reported that stellar coronae contain both hot ($10^6$ K) and
cool ($10^4$ K) plasma. In particular, time-series spectroscopy of stars
with a wide range of masses has shown periodic high-velocity
absorption and emission in Balmer lines such as H$\alpha$, interpreted
as long-lived, corotating clumps of cool plasma
\cite{CollierCameron1989,CollierCameron1992,Barnes2000,Donati2000,Dunstone2006,Skelly2008,Leitzinger2016,Cang2021}.
Such clumps are thought to be forced into corotation by the magnetic
field, and the exact geometry of where the plasma can accumulate is
dictated by the field's topology.  For instance, a magnetic dipole field
tilted with respect to the stellar spin axis
yields accumulations in a warped torus geometry
\cite{Townsend2005}, whereas in the limit of a single strong field
line, accumulation occurs at the line's apex, furthest from the star
\cite{Waugh2022}.  To date, none of these spectroscopic variables have
shown any photometric anomalies \cite{Bouma2024}, leaving open the
issue of whether they are related to CPVs.

In this study, we present the first spectroscopic detection of
corotating clumps of cool plasma around a CPV.  We identified
TIC~141146667 in previous work \cite{Bouma2024} by searching the TESS
two-minute data \cite{Ricker2015} for stars showing highly structured,
periodic light curves.  We selected the star for spectroscopy because
its brightness and rotation rate enabled an efficient search for 
variability in its line profiles.  We observed the star for five hours on
17~February~2024 (UT) using the High Resolution Echelle Spectrometer
(HIRES \cite{vogt_hires_1994}) on the 10\,m Keck I telescope.  TESS
observed the star over a four week interval from 30~January~2024 to
26~February~2024 with a duty cycle of 77\%.  During the HIRES
observations, scattered light from the Earth caused a gap in the TESS
photometry; TESS data collection resumed 12 hours (three rotation
cycles) after the spectra were acquired.  Extended Data
Figure~\ref{fig:fulllc} shows the detailed photometric behavior of the
star near the epoch of observation; the general photometric morphology
is similar before and after the data gap.

@@

% ** keck/hires: emission beyond v/veq>1
The spectroscopy shows emission beyond the star's equatorial velocity
($v_{\rm eq}$=130\,\kms).  There are at least two distinct emission
components, separated by half a cycle in phase.  The inner component at
lower velocities has clearer sinusoidal behaviour in time and is
double-peaked, with peak semi-amplitudes of 2.07\,$v_{\rm eq}$ and
2.88\,$v_{\rm eq}$ (see \S~\ref{subsec:halpha}).  There is significant
non-periodic variability in the emissivity of this double-peaked
component: the flux excess begins with an amplitude 70\% that of the
continuum early in the observation sequence, and diminishes to 10\% by
its end.  The higher-velocity component 180$^\circ$ opposite in phase is
detected only from $\phi$=0.2-1.0.  From $\phi$=0.2-0.5, this outer
component appears connected to the star in velocity space.  While its
peak semi-amplitude of 3.88\,$v_{\rm eq}$ is achieved at both
$\phi$=0.25 and 0.75, its amplitude similarly decreases from a 60\%
excess over the continuum at the beginning of the observation sequence
to a 10\% excess by its end.  The apparent period for all three
spectroscopic emission components is consistent with the photometric
\periodhr\ hour period.

These sinusoidal emission features require circumstellar clumps of
partially-ionized hydrogen to corotate with the star.  Based on the
observed sinusoidal periods and peak velocities, this material's motion
is not controlled by gravitational attraction to the star; it is more
readily explained as a partially-ionized plasma being dragged along with
a rigidly rotating stellar magnetic field.  The velocity semi-amplitude
of the sinusoids gives the mean distance of each clump from the stellar
surface: 2.07\,$R_\star$ and 2.88\,$R_\star$ for the inner clumps, and
3.88\,$R_\star$ for the outer clump.   These clumps transit in front of
the star when passing from negative to positive velocity.  The innermost
clump transits over $\approx$15\% of the period, at $\phi$=0 and
$\phi$=1.  This spectroscopic transit happens during the latter half of
the complex photometric feature in the TESS data.

% * keck/hires: behavior at v/veq<1
The H$\alpha$ line core is more complex.  At $|\Delta v / v_{\rm
eq}|<1$, most observed H$\alpha$ photons come from the star's
chromosphere; circumstellar material might also modulate the line
profile.  Figure~\ref{fig:spec}e suggests line core variability caused
by both bright and dark regions on the star's surface, superposed with
smaller-amplitude variability from the transiting circumstellar
material.  For instance, from $\phi$=0-0.3, the double-peaked emission
feature is visible when viewed both on and off-limb; this feature is
circumstellar in origin.  However, the large emission feature that
crosses the star from $\phi$=0.4-0.9 emits at an amplitude greater than
that observed from the circumstellar components, and it crosses the
stellar velocity surface at a speed that suggests it instead originates
from a chromospherically bright region on the star's surface.
Similarly, from $\phi$=0.6-1.15 a 20\% deep absorption feature slowly
crosses the H$\alpha$ line profile.  This feature suggests either a
chromospherically dark region (e.g.~a starspot group) crossing the
star's surface, or an azimuthally extended absorptive component to the
circumstellar material which is not visible in emission.  Although the
origin of the other bright and dark streaks passing across the line core
are similarly ambiguous, a final exercise to quantify the behavior of the
line core is shown in Figure~\ref{fig:spec}d, where $f_{\rm H\alpha\
core}$ denotes the summed flux at $|\Delta v / v_{\rm eq}|<1$.  This
panel shows that changes in the line core flux correlate with the
broadband variability throughout most of the light curve, except near
$\phi$$\approx$0.5, corresponding to the transit of the 3.9\,$R_\star$
clump and the occultation of the lower-velocity clump.

\begin{table*}
\small
\setlength{\tabcolsep}{2pt}
\caption{Selected system parameters for TIC~141146667}
\label{tab:params}
\begin{tabular}{llcc}
\hline \hline
Parameter & Description & Value & Source\\
%
$T_{\rm eff}$\dotfill                   & Effective Temperature (K) \hspace{9pt}\dotfill                 & 2972 $\pm$ 40    & 1 \\
%
$R_\star$\dotfill                       & Stellar radius ($R_\odot$)\dotfill                             & 0.42$\pm$0.02    & 1 \\
%
Age                                     & Stellar age range (Myr)\dotfill                                & 35-150           & 2 \\
%
$M_\star$\dotfill                       & Stellar mass ($M_\odot$)\dotfill                               & 0.22$\pm$0.02    & 3 \\
%
$\gamma$\dotfill                        & Systemic radial velocity (\kms)\dotfill                        & 0.61 $\pm$ 1.47  & 4 \\
%
SpT\dotfill                      & Spectral Type\dotfill                                          & M5.5Ve           & 4 \\
%
$P_{\rm rot}$\dotfill                   & Photometric rotation period (hr)\dotfill                       & $3.930\pm 0.001$ & 5 \\
%
$v_{\rm eq}$\dotfill		                & Equatorial velocity \dotfill                                   &  130$\pm$4       & 6 \\
                                        & \hspace{3pt} ($2\pi R_\star/P_{\rm rot}$) (\kms)	             &                      \\
%
$v_{\rm eq}\sin{i_\star}$\dotfill		    & Projected rotational\dotfill                                   &  138$\pm$8       & 4 \\
                                        & \hspace{3pt} velocity (\kms)	                                 &                      \\
%
$v_{\rm break}$\dotfill		              & Breakup velocity \dotfill                                      &  316$\pm$16      & 6 \\
                                        & \hspace{3pt} ($G M_\star / R_\star$)$^{1/2}$ (\kms)	           &                      \\
%
$i_\star$\dotfill                       & Stellar inclination\dotfill                                    & 	$>$63           & 4 \\
                                        & \hspace{3pt}  2$\sigma$ lower limit (deg)	                     &                      \\
%
$d$\dotfill                             & Distance (pc)\dotfill                                          & $57.54 \pm 0.09$ & 7 \\
%
$R_{\rm c}$\dotfill		                  & Keplerian corotation\dotfill                                   & $1.82 \pm 0.10$  & 6 \\
                                        & \hspace{3pt} radius ($R_\star$)	                               &                      \\
%
$a_0$\dotfill                           & Mean inner clump (0)\hspace{9pt}\dotfill           &  2.07$\pm$0.04   & 4 \\
                                        & \hspace{3pt} orbital radius ($R_\star$)	                       &                      \\
%
$a_1$\dotfill                           & Mean inner clump (1)\hspace{9pt}\dotfill           &  2.88$\pm$0.10   & 4 \\
                                        & \hspace{3pt} orbital radius ($R_\star$)	                       &                      \\
%
$a_2$\dotfill                           & Mean outer clump\hspace{9pt}\dotfill           &  3.88$\pm$0.25   & 4 \\
                                        & \hspace{3pt} orbital radius ($R_\star$)	                       &                      \\
%
$\langle$EW$_{\rm H\alpha}$$\rangle$    & Time-averaged H$\alpha$ line core                              &  7.2 $\pm$ 0.2   & 4 \\ 
                                        & \hspace{3pt} equivalent width (\AA)	                           &                      \\
\begin{flushleft}
\footnotesize{ \textsc{NOTE}---
Provenances are:
1: SED fit \cite{Bouma2024}.
2: Gaia DR3 photometry shows the star is on the pre-main sequence,
   while the spectrum lacks lithium (see \S~\ref{subsec:stparams}).
3: PARSEC v1.2S \cite{Chen2014}.
4: Keck/HIRES (see \S~\ref{subsec:halpha}).
5: TESS light curve.
6: Derived quantity.
7: Gaia DR3 geometric \cite{GaiaDR3}.
}
\end{flushleft}
\vspace{-0.5cm}
\end{table*}

\section{Discussion}

Spectra of magnetically-active, rapidly rotating stars with a wide
range of masses have been previously observed to exhibit both
sinusoidal time-varying Balmer emission
\cite{Donati2000,Townsend2005,Dunstone2006,Skelly2008} and transient
absorption features
\cite{CollierCameron1989,CollierCameron1992,Cang2020}, similar to
Figure~\ref{fig:spec}.  No such stars were previously known to show
complex light curves \cite{Bouma2024}.   One interpretation for the
spectroscopic variability comes from an analogy to quiescent
solar prominences, which are cool condensations of plasma in the solar
corona that can last days to weeks \cite{VialEngvold2015}.  These
condensations fall back to the Sun's photosphere because gravity is
stronger than any magnetic tension or centrifugal force capable of
sustaining them.  However for stars with magnetospheric radii $R_{\rm
m}$ that exceed their corotation radii $R_{\rm c}$, the effective
potential experienced by a charged particle can have a local minimum
outside $R_{\rm c}$, enabling the material to accumulate in a
centrifugally-supported magnetosphere
\cite{Petit2013,Daley-Yates2024}.  Generally speaking, such material
need neither transit nor be opaque in broadband optical light.  In
fact, Figure~\ref{fig:spec} suggests that the transits of the
H$\alpha$ emitting material cannot explain the entirety of the complex
photometric modulation;  the spectroscopic transits happen too fast, and
at orbital phases that only partially overlap with the photometric
modulations.

Nonetheless, our Keck/HIRES observations are the first reported
time-series spectra of a CPV, and they show for the first time that
corotating circumstellar plasma clumps exist around at least one such
star.  While starspots likely do contribute smooth photometric
variability signals, a ``starspot-only'' scenario for the CPVs
\cite{Koen2021} seems to be incompatible with our observations.
Scenarios in which the circumstellar material is made only of dust are
similarly ruled out.  While starspots and dust may both be present,
the observed H$\alpha$ emission requires clumps of plasma with a
significant population of hydrogen atoms in the $n$=3 excited state,
with most H$\alpha$ emission coming from a region half the size of the
star (see \S~\ref{subsec:halpha}).  The observed H$\alpha$ luminosity
suggests
characteristic densities and masses for the gaseous component of these
clumps of $n_{\rm H} \sim 10^{11}$-$10^{12}$\,cm$^{-3}$ and $M_{\rm
gas} \sim 10^{17}$\,g (see \S~\ref{subsec:gas}).  Dust, if present, is
independently constrained to have a total mass $10^{15}\,{\rm g} <
M_{\rm dust} < 10^{17}\,{\rm g}$ (see \S~\ref{subsec:dust}).  This
circumstellar material could originate either from the star or from an
external source.  Plausible external sources include an undetected old
disk, comets, or a close-in exoplanet.  This latter scenario would
make CPVs extrasolar analogs of the Jupiter-Io plasma torus
(e.g.~Ref.~\cite{Bagenal1981}).

The other potential analog for the CPVs are the $\sigma$~Ori~E
variables, a rare subset of B stars with radiatively-driven winds
which accumulate into warped plasma tori
\cite{Townsend2005,Townsend2008}.  These tori tend to have dense
antipodal accumulations of plasma sculpted by tilted-dipole magnetic
fields, and the transits of these clumps produce broadband photometric
variability through bound-free scattering \cite{Townsend2005} and
Thomson scattering \cite{Berry2022}.  The plasma accumulates at
antipodes 180$^\circ$ apart because the deepest local minima in the
effective potential exist along the line of intersection between the
rotational and magnetic equators (see Equation~22 of
\cite{Townsend2005}).  For $\sigma$~Ori~E and almost all of its
analogs, the result is ``simple'' light curves that resemble those of
eclipsing binaries, and time-dynamic H$\alpha$ spectra similar to
those in Figure~\ref{fig:spec} \cite{Townsend2005,Townsend2008}.  The
two known exceptions, HD~37776 and HD~64740, show complex light curves
resembling CPVs \cite{Mikulasek2020,Bouma2024} and have
spectropolarimetric magnetic field maps indicating strong
contributions from higher-order magnetic moments
\cite{Kochukhov2011,Shultz2018}.  There are two implications: first, the
photometric complexity of CPVs may be a direct consequence of magnetic
fields with highly multipolar contributions.  Second, despite this
complexity, the observation from Figure~\ref{fig:spec} of two emission
clumps separated 180$^\circ$ in phase is consistent with the expected
emission from a warped plasma torus.

@@

The material's composition -- either pure plasma or dusty plasma --
can be clarified by time-series optical and infrared
spectrophotometry.  While observations of CPVs in the optical suggest
a chromaticity consistent with dust
\cite{Tanimoto2020,Gunther2022,Koen2023}, a gray opacity source such
as electron scattering in a plasma transiting over a spotted
background star might also produce chromatic features
\cite{Rackham2018}.  The composition and size distribution of any dust
that is present could be determined by measuring the extinction curve
for a sample of CPVs from 1-20\,$\mu$m.  Composition and size
distributions similar to debris from rocky bodies seen around white
dwarfs \cite{Reach2009} would suggest an extrinsic origin channel.
Compositions and sizes similar to the interstellar medium would
suggest that dust can condense out from M dwarf winds, similar to
processes that occur around more evolved stars \cite{Marigo2008}.

@@

%\item[Author Contributions] LGB is responsible for all aspects of this
%  work, including conceptualization, data analysis, interpretation,
%  manuscript preparation, and revision.
%
\item[Data Availability] The TESS two-minute cadence data used in this
  work are publicly available from the Mikulski Archive for Space
  Telescopes (MAST; \url{https://mast.stsci.edu}). The Keck/HIRES
  spectra analyzed here are accessible via the Keck Observatory
  Archive (KOA; \url{https://koa.ipac.caltech.edu}).
